MassWateR: Improving quality control, analysis, and sharing of water quality data

### 2023-07-28 paper.Rmd
revisions form JC comments done

---
title: "MassWateR: Improving quality control, analysis, and sharing of water quality data"
date: "`r Sys.Date()`"
abstract: >
  The long-term protection and restoration of aquatic resources depends on accurate and precise monitoring data.  These data require systematic quality control and analysis tools for regulatory applications or routine assessments.  The \CRANpkg{MassWateR} package provides several functions that can facilitate quality control, analysis, and data submission for water quality data collected from different monitoring programs.  The tools were developed to address specific regional needs for Massachusetts, USA, but the principles can be generalized to monitoring data collected at any location.  Users can create quality control reports, perform outlier analyses, and assess various trends by season, date, and site.  Functions are also included to prepare data for submission to the Water Quality Exchange data portal, the largest water quality database in the United States.  These automated and reproducible assessments are expected to increase the amount of usable data for the evaluation of aquatic resources.  
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Marcus W. Beck
    affiliation: Tampa Bay Estuary Program
    address:
    - 263 13th Ave S
    - St. Petersburg, Florida, USA 33701
    url: https://tbep.org
    orcid: 0000-0002-4996-0059
    email:  mbeck@tbep.org
  - name: Benjamen Wetherill
    url: https://www.acasak.com/
    email: bwetherill@acasak.com
    affiliation: ACASAK Consulting
    address:
    - Boston, Massachusetts, USA
  - name: Jillian Carr
    url: https://www.mass.gov/orgs/massachusetts-bays-national-estuary-partnership
    email: Jillian.Carr@umb.edu
    affiliation: Massachusetts Bays National Estuary Partnership
    address:
    - University of Massachusetts Boston, 100 Morrissey Blvd
    - Boston, Massachusetts, USA 02125
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble: |
  \usepackage{multirow}
  \usepackage{multicol}
  \usepackage{colortbl}
  \usepackage{hhline}
  \newlength\Oldarrayrulewidth
  \newlength\Oldtabcolsep
  \usepackage{longtable}
  \usepackage{array}
  \usepackage{hyperref}
  \usepackage{float}
  \usepackage{wrapfig}\usepackage{longtable}
---

@@

Water quality measurements provide the foundation of environmental monitoring programs designed to protect or restore aquatic resources. In the United States, these programs are broadly guided by the federal Clean Water Act with the singular goal of restoring and maintaining the chemical, physical, and biological integrity of the nation's surface waters.  Similarly, the Water Framework Directive provides a framework for the protection of aquatic resources in member states of the European Union.  Numeric standards that define critical thresholds for protecting recreational, aquatic life, industrial, navigational, and consumptive uses of the resource are often established, that, if exceeded based on water quality measurements, require additional regulatory action to ensure compliance. These standards and other regulatory assessments as applied at the state-level use information from long-term monitoring datasets [@Schiff16; @Tango13], or data collected *ad hoc* from multiple assessment endpoints [@Stein09; @Behmel16; @Kumpel20], where the former is atypical for most surface water bodies.  Many state or regional institutions that assess water quality rely on decentralized data sources, often combining datasets from local watershed groups or participatory science programs rather than a single database that contains adequate coverage of all areas of interest [@Buytaert14; @Kelly22]. Use of these monitoring data in a regulatory context is not possible unless standard operating procedures are adopted and the data fulfill quality control (QC) requirements.

Monitoring data of sufficient quantity and quality are critical to ensure precise and accurate representation of environmental conditions.  A significant bottleneck in the use of monitoring data for environmental assessment of surface waters is the ability to clearly and efficiently indicate that the data fulfill applicable QC checks for regulatory applications or inclusion in a consolidated database [@Arndt22]. Common QC checks for *in situ* field measurements or concentrations measured in the laboratory may include 1) comparison of the precision between replicate samples (duplicates), 2) accuracy of a sample relative to a known concentration (spikes or instrument checks), and 3) evaluation of the measurement from an empty or blank sample (blanks) [@Wilde02].  An adequate number of QC samples must also be included in the dataset as a measure of "completeness".  These checks are often compiled in a single report for review by appropriate regulatory agencies.  For example, precision of duplicate samples for a given parameter must not vary more than 5% and at least 10% of the data should be dedicated to these checks as a measure of completeness. For local monitoring groups that lack the resources to develop robust and repeatable workflows, QC reports are often prepared manually before submitting the data for review.  This process is time-consuming and prone to errors, often limiting the amount of useful information for regulatory assessments or submitted to formal databases.

The use of R with document generation systems offered through packages like \CRANpkg{knitr} [@Xie15] and \CRANpkg{rmarkdown} [@Allaire23] can generate QC reports that follow a standard format for review by regulatory agencies.  These tools can also be used to format water quality data for submission to state or national water quality databases, such as the Water Quality Exchange (WQX) database maintained by the US Environmental Protection Agency (USEPA).  This database is the largest source of monitoring data in the United States that includes information on hydrologic conditions and chemical, physical, and biological measurements from surface waters.  Further, many environmental resource managers have the need to analyze status and trends in monitoring data and R packages such as \CRANpkg{ggplot2} [@Wickham16] and others in the \CRANpkg{tidyverse} [@Wickham19] offer useful tools to synthesize and visualize results. Integrating this functionality into a single package is expected to have wide ranging utility for anyone collecting surface water data and is likely to improve the quality and insights obtained from these data.

This paper describes the \CRANpkg{MassWateR} package developed to improve how environmental professionals perform quality control, analysis, and sharing of monitoring data for surface waters.  The regional focus of the package is for monitoring data collected in Massachusetts, USA, with QC reports submitted to the Massachusetts Department of Environmental Protection (MADEP) and data submitted to the national WQX database.  Although the initial conception of \CRANpkg{MassWateR} was to address regional needs in Massachusetts, there is nothing specific in the package that prevents its use outside of the state as the QC checks and analyses follow routine and commonly used methods for data collected elsewhere.  As such, this paper is written with emphasis on how the tools are broadly applicable to anyone interested in improving efficiency and reproducibility of QC checks, in addition to analysis of water quality data and submission to WQX as the largest source of water quality monitoring data in the US.

@@

To our knowledge, there are no existing R packages on CRAN that can be used to facilitate QC of water quality data, nor are any available that facilitate submission to existing databases.  However, there are several that can be used to retrieve and analyze data from existing sources (see the CRAN \ctv{Hydrology} Task View).  In particular, the \CRANpkg{dataRetrieval} package [@DeCicco22] has been used widely to retrieve data from the USEPA Water Quality Portal (WQP), which is the counterpart of the WQX system for accessing data submitted using the latter.  This package leverages a robust API to query existing water quality data in standardized format provided by the WQP.  As such, data retrieval using existing web services is much simpler than data submission, as data formatting requirements do not apply when retrieving data.  Developing a robust tool that can facilitate the upload of data to WQX, in addition to streamlining QC processes, would further the value of packages like \CRANpkg{dataRetrieval} by increasing the amount of data that can be accessed through the WQP.  The \CRANpkg{MassWateR} package was developed to provide this benefit.

Users can engage with \CRANpkg{MassWateR} to achieve different goals.  This design was intentional based on likely differences in needs among the user community.  Although increasing data submission and facilitating QC reporting was the primary goal, we also assumed that users may not want to do both.  That is, state institutions require QC reporting for regulatory assessment, whereas data submission to WQX may be a separate process.  Users may also simply have a need to understand trends or to summarize their data, while also wanting to extend these analyses beyond \CRANpkg{MassWateR} using additional R packages. Figure \@ref(fig:workflow) demonstrates how a user may apply the functions in \CRANpkg{MassWateR} once the required data are imported.  The functions allow a user to engage with their data several ways, including 1) screening data for quality control, 2) summarizing quality control results into a single report or separate tables, 3) creating graphics for analysis and reports to stakeholders, and 4) formatting data for upload to WQX.

```{r workflow, fig.cap = 'Workflow demonstrating how a user could engage with the \\CRANpkg{MassWateR} package.  A user can apply one to any of the four steps depending on their need.  The first step, QC screening, is often iterative as a user can modify parts of the raw data based on input checks or outliers.  The second step can be used to create a QC report for submission to a regulatory agency.  The third step can create a formatted table for WQX submission.  The fourth step is data analysis and visualization, using \\CRANpkg{MassWateR} functions and downstream analysis with additional R packages and functions.  All steps require formatted input files.  WQX: Water Quality Exchange; QC: Quality Control.', fig.alt='Workflow diagram showing four ways to engage with MassWateR', out.width='100%'}
knitr::include_graphics('figs/workflow.png')
```

No matter the user need, all data inputs to \CRANpkg{MassWateR} must follow a strict format.  Developing a workflow to accommodate data inputs from the dozens of potential users from different organizations that use different data formats would have been impractical.  As such, the only limitation to using the package is to adhere to the formatting requirements for all input files.  Several [resources](https://massbays-tech.github.io/MassWateR/RESOURCES.html) are provided on the package web page to assist users in formatting their data.  These resources included several training activities that were conducted during package development and templates demonstrating the appropriate format and rationale.

The required data files for using \CRANpkg{MassWateR} are shown in Table \@ref(tab:filreq), including the files that apply to the workflow steps in Figure \@ref(fig:workflow).  The files are each imported into R using specific `read` functions with relevant checks, explained in the next section.  These checks verify multiple requirements outlined in the template files, with informative errors or warnings returned to the console to prompt the user on the required action to remedy a formatting issue.  The largest input file required for all parts of the workflow in Figure \@ref(fig:workflow) is the results file.  This file includes all water quality monitoring data to be used with the package and the recommendation is that the results data include a year of sampling as reporting typically follows an annual cycle. As such, the formatting requirements for the results file are the most burdensome for potential users and additional functions are available to assist in this effort.

@@

The primary task of the `read` functions is to ensure all imported files follow the required format for the package.  Excel files are the expected format for all inputs and the `read` functions use the `read_excel` function from the \CRANpkg{readxl} package [@Wickham23]. The `read` functions do very little other than import the file - once the file is imported it is immediately passed to one of the relevant `check` functions inside the `read` function.  There are several checks for each type of input file, with the number of checks increasing based on the complexity of the input file.  Each check is printed to the R console on completion, whereas an error is returned at the first instance of a failed check, at which point the function is exited.  The error will typically indicate which parts of the input file need to be changed to rectify the issue, often indicating a specific cell in the Excel file that requires attention. As such, the workflow is intended to be iterative, where a user imports a file, receives an error, manually changes the input file in Excel, then imports the data again until all checks pass (Figure \@ref(fig:importflow)).  Again, this design was intentional as many monitoring agencies and groups collect data differently and a standard input format for the package was the best option to accommodate all potential users.  This may also encourage future standardization among monitoring groups for how data are maintained to ease formatting challenges to using \CRANpkg{MassWateR}. A user only needs to format their data once to use the package.

@@

The `readMWRresultsview()` function is also available to assist with troubleshooting formatting issues for the results file.  This function exports an Excel file that shows the unique values that are found in each column to allow a user to quickly see potential incorrect entries.  The output is similar to running the following:

@@

After all file format checks are fixed, a standard approach for using \CRANpkg{MassWateR} is to import all required files and save them as a list of named data frame objects that can be used by nearly all the package functions.  This prevents the need to identify which input datasets are needed for each function, although the latter approach could used because arguments for individual input files are also provided in all functions.  In the latter case, a path or data object can be used as input for each file. For the former approach, the beginning of a script for using the package could appear as follows.  Example files included with the package are imported, whereas a user will specify paths to their own files.

@@

Checking the results file for outliers is an additional and essential step of QC screening. Outliers can be caused by several reasons, including data entry errors, laboratory processing errors, or anomalous environmental events.  Identifying and potentially removing or correcting outliers is a critical part of quality control.  The functions in \CRANpkg{MassWateR} simply identify potential outliers to provide an opportunity for users to address these values.  The decision on how to address outliers remains with the user and no automated tools are provided in the package to correct or remove outliers.

@@

A user can also evaluate outliers for every water quality parameter in the results file.  The `anlzMWRoutlierall()` function creates a Word document with images of boxplots for every parameter.  Images for each parameter can also be created as standalone files.  The `param` argument does not need to be specified because all parameters are identified in the results file and processed accordingly.  The following shows how to use the function by creating a single Word file or individual image files in the working directory.

@@

```{r, echo = T, eval = F}
qcMWRreview(fset = fsetls, output_dir = getwd())
#> Report created successfully! File located at /tmp/RtmpUzzrbC/qcreview.docx
```

@@

The cells in red showed that four of the checks for three of the parameters did not include a sufficient amount of QC records that passed the relevant accuracy checks.  All other parameters and checks in green had enough QC checks to satisfy the DQOs.  Empty cells include checks where no QC records were available in the results file, e.g., no lab duplicates for dissolved oxygen.  The empty cells are typically for QC checks that do not readily apply to a parameter.  For example, dissolved oxygen is measured in the field with monitoring equipment, such that lab QC checks are not relevant.

@@

The example shows that all but one field blank passed the checks, where most were below the detection limit (BDL) of the laboratory equipment.  Only one sample taken on July 17th for total phosphorus (TP) was at or above the threshold.

Additional functions for creating QC tables include `tabMWRfre()` for assessing if a sufficient number of QC samples were performed and `tabMWRcom()` for assessing completeness checks that compare the number of regular samples (field measurements or lab samples) to the number of qualified samples.  As with all the QC functions, these can be used individually within an R session to quickly view QC results or the `qcMWRreview()` function can be used to create a complete report that combines results from the individual functions.  The summarized report can then be submitted to an appropriate regulatory agency for review to ensure that any submitted datasets fulfill appropriate data quality objectives.

@@

The analysis functions in \CRANpkg{MassWateR} provide a simple approach to quickly evaluate data in the results file.  Although several base R functions and supporting packages can be used to develop custom assessments, it was recognized that many users will not be comfortable developing their own visualization routines.  As such, the analysis functions were designed for a rapid overview of the results that required minimal decisions by the user to create the output.  Several default settings described below allow further customization of the output as desired by the user. The four primary analysis functions can be used to analyze seasonal trends, trends by date, data by site, and spatially using maps.

@@

The y-axis scaling of the plot is also determined automatically from the data inputs.  The DQO file for accuracy includes information on the distribution of each parameter, i.e., parameters with "log" in any of the columns are plotted on log10-scale, otherwise arithmetic.  This behavior is controlled by the `yscl` argument, where the default is `"auto"` that indicates information on scaling is obtained automatically from the DQO file.  Setting `yscl = "linear"` or `yscl = "log"` will set the axis as linear or log10-scale, respectively, regardless of the information in the DQO file.

@@

The `anlzMWRsite()` function can be used to view data organized by site. This function summarizes results for a single parameter using boxplots or barplots separately for each site on the x-axis. Boxplots or barplots can also include jittered points of the observations on top or only the jittered points can be shown, as for the `anlzMWRseason()` function. Results at each site can also be grouped by the `Result Attribute` column in the results file, where a user can enter arbitrary grouping criteria for samples. For example, sites can be grouped by wet or dry conditions if this information is included in the results file.  Individual sites to plot can also be specified with the `site` argument, where the default is to plot all sites.

@@

Finally, maps of results at each site can be created with the `anlzMWRmap()` function.  Parameter values at each site are summarized with colors indicating relative values and the maps can include various spatial information for context. For example, hydrologic lines and waterbodies from the National Hydrograpy Dataset (NHD) can be shown with varying level of detail and base maps can be included from the \CRANpkg{ggmap} package [@Kahle13].  NHD data were included to provide more specific information on hydrologic features of interest because of insufficient detail provided by standard basemaps.  These datasets are available from an external source and clipped to an approximate bounding box for the selected stations.  Currently, only flowlines and waterbodies for Massachusetts are included, although examples are provided in a package [vignette](https://massbays-tech.github.io/MassWateR/articles/modifying.html#modifying-masswater-plots) that demonstrate how to include custom shapefiles in the map.  The below example shows mean dissolved oxygen concentrations using the medium level of detail for the NHD maps (`addwater = "medium"`).

@@

The last part of the \CRANpkg{MassWateR} workflow is preparation of data for submission to WQX.  A single function, `tabMWRwqx()`, is provided to format data inputs using a [template](https://massbays-tech.github.io/MassWateR/RESOURCES.html) designed for data submission.  This function will export a single Excel workbook with three sheets, named "Project", "Locations", and "Results", all of which are required for data submission. The output is populated with as much content as possible based on information in the input files and the user can manually adjust the workbook prior to submission. All required columns are present, but individual rows must be verified for completeness and accuracy by the user before uploading the data.

The workbook can be created as follows by including the required files and specifying an output directory where the Excel file is saved. Once the function is done running, a message indicating success and where the file is located is returned.  Submitting the data to WQX simply requires copying/pasting the content in the output file into the data portal.

@@

Additional templates and instructions for data submission are available on the [package website](https://massbays-tech.github.io/MassWateR/index.html).  It is assumed that users are familiar with the WQX data submission portal if they are already using the package.  \CRANpkg{MassWateR} eliminates the need to format the data by hand and it is expected to increase the amount of data made available on the WQP (via WQX upload) as the user base for the package increases.

>>>>

---
title: "MassWateR: Improving quality control, analysis, and sharing of water quality data"
date: "`r Sys.Date()`"
abstract: >
  The long-term protection and restoration of aquatic resources depends on robust monitoring data; data that require systematic quality control and analysis tools. The \CRANpkg{MassWateR} package facilitates quality control, analysis, and data sharing for discrete surface water quality data collected by monitoring programs of various size and technical capacity. The tools were developed to address regional needs for programs in Massachusetts, USA, but the principles and outputs can be applicable to monitoring data collected anywhere.  Users can create quality control reports, perform outlier analyses, and assess trends by season, date, and site for more than 40 parameters. Users can also prepare data for submission to the USEPA Water Quality Exchange, thus sharing data to the largest water quality database in the United States. The automated and reproducible workflow offered by \CRANpkg{MassWateR} is expected to increase the quantity and quality of publicly available data for the evaluation of aquatic resources.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Marcus W. Beck
    affiliation: Tampa Bay Estuary Program
    address:
    - 263 13th Ave S
    - St. Petersburg, Florida, USA 33701
    url: https://tbep.org
    orcid: 0000-0002-4996-0059
    email:  mbeck@tbep.org
  - name: Benjamen Wetherill
    url: https://www.acasak.com/
    email: bwetherill@acasak.com
    affiliation: ACASAK Consulting
    address:
    - Boston, Massachusetts, USA
  - name: Jillian Carr
    url: https://www.mass.gov/orgs/massachusetts-bays-national-estuary-partnership
    email: Jillian.Carr@umb.edu
    affiliation: Massachusetts Bays National Estuary Partnership
    address:
    - University of Massachusetts Boston, 100 Morrissey Blvd
    - Boston, Massachusetts, USA 02125
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble: |
  \usepackage{multirow}
  \usepackage{multicol}
  \usepackage{colortbl}
  \usepackage{hhline}
  \newlength\Oldarrayrulewidth
  \newlength\Oldtabcolsep
  \usepackage{longtable}
  \usepackage{array}
  \usepackage{hyperref}
  \usepackage{float}
  \usepackage{wrapfig}\usepackage{longtable}
---

@@

Water quality measurements provide the foundation of environmental monitoring programs designed to protect or restore aquatic resources. In the United States, these programs are broadly guided by the federal Clean Water Act with the singular goal of restoring and maintaining the chemical, physical, and biological integrity of the nation's surface waters.  Similarly, the Water Framework Directive provides a similar legislative foundation for the protection of aquatic resources in member states of the European Union.  Numeric standards that define critical thresholds for protecting recreational, aquatic life, industrial, navigational, and consumptive uses of the resource are often established by government agencies, such that exceedances idenfitied from water quality measurements require additional regulatory action to ensure compliance. These standards and other regulatory assessments as applied at the state-level use information from long-term monitoring datasets [@Schiff16; @Tango13], or data collected *ad hoc* from multiple assessment endpoints [@Stein09; @Behmel16; @Kumpel20], where the former is atypical for most surface water bodies.  Many state or regional institutions that assess water quality rely on decentralized data sources, often combining datasets from local watershed groups or participatory science programs rather than a single database that contains adequate coverage of all areas of interest [@Buytaert14; @Kelly22]. Use of these monitoring data in a regulatory context is not possible unless standard operating procedures are adopted and the data fulfill quality control (QC) requirements.

Monitoring data of sufficient quantity and quality are critical to ensure precise and accurate representation of environmental conditions.  A significant bottleneck in the use of monitoring data for environmental assessment of surface waters is the ability to clearly and efficiently indicate that the data fulfill applicable QC checks for regulatory applications or inclusion in a consolidated database [@Arndt22]. Common QC checks for *in situ* field measurements or concentrations measured in the laboratory may include 1) comparison of the precision between replicate samples (duplicates), 2) accuracy of a sample relative to a known concentration (spikes or instrument checks), and 3) evaluation of the measurement from an empty or blank sample (blanks) [@Wilde02].  An adequate number of QC samples must also be included in the dataset as a measure of "completeness".  These checks are often compiled in a single report for review by appropriate regulatory agencies.  For example, precision of duplicate samples for a given parameter must not vary more than 5% and at least 10% of the data should be dedicated to these checks as a measure of completeness. For local monitoring groups that lack the resources to develop robust and repeatable workflows, QC reports are often prepared manually before submitting the data for potential use by agencies.  This process is time-consuming and prone to errors, often limiting the amount of useful information for regulatory assessments or submitted to formal databases.  As an example, dozens of community-driven environmental organizations in Massachusetts, USA, collect water quality data, yet government agencies are only able to access and utilize a fraction of those datasets. The Massachusetts Bays National Estuary Partnership funded the development of \CRANpkg{MassWateR} described herein to build capacity among water monitoring programs to generate and share high quality data. This capacity need is not specific to Massachusetts, thus, the tools were developed with transferability in mind.

The use of R with document generation systems offered through packages like \CRANpkg{knitr} [@Xie15] and \CRANpkg{rmarkdown} [@Allaire23] can generate QC reports that follow a standard format for review by regulatory agencies.  These tools can also be used to format water quality data for submission to state or national water quality databases, such as the Water Quality Portal (WQP) database maintained by the US Environmental Protection Agency (USEPA) and contributed to via the Water Quality eXchange (WQX).  This database is the largest source of monitoring data in the United States that includes information on hydrologic conditions and chemical, physical, and biological measurements from surface waters.  Further, many environmental resource managers have the need to analyze status and trends in monitoring data and R packages such as \CRANpkg{ggplot2} [@Wickham16] and others in the \CRANpkg{tidyverse} [@Wickham19] offer useful tools to synthesize and visualize results. Integrating this functionality into a single package is expected to have wide ranging utility for anyone collecting surface water data and is likely to improve the quality and insights obtained from these data.

This paper describes the \CRANpkg{MassWateR} package developed to improve how environmental professionals perform quality control, analysis, and sharing of monitoring data for surface waters.  The regional focus of the package is for monitoring data collected in Massachusetts, USA, with QC reports submitted to the Massachusetts Department of Environmental Protection (MADEP) and data submitted to the national WQP database.  Although the initial conception of \CRANpkg{MassWateR} was to address regional needs in Massachusetts, there is nothing specific in the package that prevents its use outside of the state as the QC checks and analyses follow routine and commonly used methods for data collected elsewhere.  As such, this paper is written with emphasis on how the tools are broadly applicable to anyone interested in improving efficiency and reproducibility of QC checks, in addition to analysis of water quality data and submission to WQX as the largest source of water quality monitoring data in the US.

@@

To our knowledge, there are no existing R packages on CRAN that can be used to facilitate QC of water quality data, nor are any available that facilitate submission to existing databases.  However, there are several that can be used to retrieve and analyze data from existing sources (see the CRAN \ctv{Hydrology} Task View).  In particular, the \CRANpkg{dataRetrieval} package [@DeCicco22] has been used widely to retrieve data from the WQP, which is the interface for accessing data submitted using WQX.  This package leverages a robust API to query existing water quality data in standardized format provided by the WQP.  As such, data retrieval using existing web services is much simpler than data submission, as data formatting requirements do not apply when retrieving data.  Developing a robust tool that can facilitate the upload of data to WQX, in addition to streamlining QC processes, would further the value of packages like \CRANpkg{dataRetrieval} by increasing the amount of data that can be accessed through the WQP.  The \CRANpkg{MassWateR} package was developed to provide this benefit.

Users can engage with \CRANpkg{MassWateR} to achieve different goals.  This design was intentional based on likely differences in needs among the user community.  Although increasing data submission and facilitating QC reporting was the primary goal, we also assumed that users may not want to do both.  That is, state institutions require QC reporting for regulatory assessments, whereas data submission to WQX may be a separate process.  Users may also simply have a need to explore trends or to summarize their data, while also wanting to extend these analyses beyond \CRANpkg{MassWateR} using additional R packages. Figure \@ref(fig:workflow) demonstrates how a user may apply the functions in \CRANpkg{MassWateR} once the required data are imported.  The functions allow a user to engage with their data several ways, including 1) screening data for quality control, 2) summarizing quality control results into a single report or separate tables, 3) creating graphics for analysis and reports to stakeholders, and 4) formatting data for upload to WQX. A user can apply one to any of the four steps depending on their need.  The first step, QC screening, is often iterative as a user can modify parts of the raw data based on input checks or outliers.  The second step can be used to create a QC report for submission to a regulatory agency.  The third step can create a formatted table for WQX submission.  The fourth step is data analysis and visualization, using \CRANpkg{MassWateR} functions and downstream analysis with additional R packages and functions.  All steps require formatted input files.

```{r workflow, fig.cap = 'Workflow demonstrating how a user could engage with the \\CRANpkg{MassWateR} package. WQX: Water Quality Exchange; QC: Quality Control.', fig.alt='Workflow diagram showing four ways to engage with MassWateR', out.width='100%'}
knitr::include_graphics('figs/workflow.png')
```

No matter the user need, all data inputs to \CRANpkg{MassWateR} must follow a strict format.  Developing a workflow to accommodate data inputs from the dozens of potential users from different organizations that use different data formats would have been impractical.  As such, the primary limitation to using the package is to adhere to the formatting requirements for all input files.  Several [resources](https://massbays-tech.github.io/MassWateR/RESOURCES.html) are provided on the package web page to assist users in formatting their data.  These resources included several training activities that were conducted during package development and templates demonstrating the appropriate format and rationale.  The trainings, \CRANpkg{pkdown} [website](https://massbays-tech.github.io/MassWateR/) [@Wickham22], and [Community of Practice forum](https://massbays.discourse.group/login) were implemented so that learning R was not a significant limitation for using the package.

The required data files for using \CRANpkg{MassWateR} are shown in Table \@ref(tab:filreq), including the files that apply to the workflow steps in Figure \@ref(fig:workflow).  The files are each imported into R using specific `read` functions with relevant checks, explained in the next section.  These checks verify multiple requirements outlined in the template files, with informative errors or warnings returned to the console to prompt the user on the required action to remedy a formatting issue.  The largest input file required for all parts of the workflow in Figure \@ref(fig:workflow) is the results file.  This file includes all water quality monitoring data to be used with the package and the recommendation is that the results data include a year of sampling as reporting typically follows an annual cycle. The results file design also considered WQX requirements to facilitate data submission. As such, the formatting requirements for the results file are the most burdensome for potential users and additional functions are available to assist in this effort.

@@

The primary task of the `read` functions is to ensure all imported files follow the required format for the package.  Excel files are the expected format for all inputs and the `read` functions use the `read_excel` function from the \CRANpkg{readxl} package [@Wickham23]. The `read` functions do very little other than import the file - once the file is imported it is immediately passed to one of the relevant `check` functions inside the `read` function.  There are several checks for each type of input file, with the number of checks increasing based on the complexity of the input file.  Each check is printed to the R console on completion, whereas an error is returned at the first instance of a failed check, at which point the function is exited.  The error will typically indicate which parts of the input file need to be changed to rectify the issue, often indicating a specific cell in the Excel file that requires attention. As such, the workflow is intended to be iterative, where a user imports a file, receives an error, manually changes the input file in Excel, then imports the data again until all checks pass (Figure \@ref(fig:importflow)).  Again, this design was intentional as many monitoring agencies and groups organize data differently and a standard input format for the package was the best option to accommodate all potential users.  This may also encourage future standardization among monitoring groups for how data are maintained to ease formatting challenges to using \CRANpkg{MassWateR}. A user only needs to format their data once to use the package.

@@

The `readMWRresultsview()` function is also available to assist with troubleshooting formatting issues for the results file.  This function exports an Excel file that shows the unique values that are found in each column to allow a user to quickly see potential incorrect entries or typos.  The output is similar to running the following:

@@

After all file format checks are completed and errors fixed, a standard approach for using \CRANpkg{MassWateR} is to import all required files and save them as a list of named data frame objects that can be used by nearly all the package functions.  This prevents the need to identify which input datasets are needed for each function, although the latter approach could be used because arguments for individual input files are also provided in all functions.  In the latter case, a path or data object can be used as input for each file. For the former approach, the beginning of a script for using the package could appear as follows.  Example files included with the package are imported, whereas a user will specify paths to their own files.

@@

Checking the results file for outliers is an additional and essential step of QC screening. Outliers can have many causes, including data entry errors, laboratory processing errors, or anomalous environmental events.  Identifying and potentially removing or correcting outliers is a critical part of quality control.  The functions in \CRANpkg{MassWateR} simply identify potential outliers to provide an opportunity for users to address these values.  The decision on how to address outliers remains with the user and no automated tools are provided in the package to correct or remove outliers.

@@

A user can also evaluate outliers for every water quality parameter in the results file.  The `anlzMWRoutlierall()` function creates a Word document with images of boxplots for every parameter.  Images for each parameter can also be created as standalone files.  Here, the `param` argument does not need to be specified because all parameters are identified in the results file and processed accordingly.  The following shows how to use the function to create a single Word file or individual image files in the working directory.

@@

```{r, echo = T, eval = F}
qcMWRreview(fset = fsetls, output_dir = getwd())
```
```
#> Report created successfully! File located at /tmp/RtmpUzzrbC/qcreview.docx
```

@@

The cells in red showed that four of the QC checks for three of the parameters did not meet the accuracy DQOs.  All other parameters and checks in green had enough QC checks to satisfy the DQOs (note that the table colors are color-blind friendly).  Empty cells include checks where no QC records were available in the results file, e.g., no lab duplicates for dissolved oxygen.  The empty cells are typically for QC checks that do not readily apply to a parameter.  For example, dissolved oxygen is measured in the field with monitoring equipment, such that lab QC checks are not relevant.

@@

The example shows that all but one field blank passed the checks, where most were below the detection limit (BDL) of the laboratory equipment.  Only one sample taken on July 17th for total phosphorus (TP) was at or above the threshold (i.e., was a "MISS" for the DQO target).

Additional functions for creating QC tables include `tabMWRfre()` for assessing if QC samples were conducted at the targeted frequency and `tabMWRcom()` for assessing completeness checks that compare the number of regular samples (field measurements or lab samples) to the number of qualified samples, i.e., a measure of the proportion of the data that is usable.  As with all the QC functions, these can be used individually within an R session to quickly view QC results or the `qcMWRreview()` function can be used to create a complete report that combines outputs from the individual functions.  The summarized report can then be submitted to an appropriate regulatory agency for review to ensure that any submitted datasets fulfill appropriate data quality objectives.  The report provides space for the user to add narrative context and notes prior to submission.

@@

The analysis functions in \CRANpkg{MassWateR} provide a streamlined approach to quickly evaluate data in the results file.  Although several base R functions and supporting packages can be used to develop custom assessments, it was recognized that many users will not be comfortable developing their own visualization routines.  As such, the analysis functions were designed for a rapid overview of the results that required minimal decisions by the user to create the output.  Several default settings described below allow further customization of the output as desired by the user. The four primary analysis functions can be used to analyze seasonal trends, trends by date, data by site, and spatially using maps.

@@

The y-axis scaling of the plot is also determined automatically from the data inputs.  The DQO file for accuracy includes information on the distribution of each parameter, i.e., parameters with "log" in any of the columns are plotted on log10-scale, otherwise arithmetic.  This behavior is controlled by the `yscl` argument, where the default is `"auto"` which indicates information on scaling is obtained automatically from the DQO file.  Setting `yscl = "linear"` or `yscl = "log"` will set the axis as linear or log10-scale, respectively, regardless of the information in the DQO file.

@@

The `anlzMWRsite()` function can be used to view data organized by site. This function summarizes results for a single parameter using boxplots or barplots separately for each site on the x-axis. Boxplots or barplots can also include overlaid jittered points of the observations or only the jittered points can be shown, as for the `anlzMWRseason()` function. Results at each site can also be grouped by the `Result Attribute` column in the results file, where a user can enter arbitrary grouping criteria for samples. For example, sites can be grouped by wet or dry conditions if this information is included in the results file to evaluate the effects of precipitation on water chemistry or pollutants.  Individual sites to plot can also be specified with the `site` argument, where the default is to plot all sites.

@@

Finally, maps of results at each site can be created with the `anlzMWRmap()` function.  Parameter values at each site are summarized with colors indicating relative values and the maps can include various spatial information for context. For example, hydrologic lines and waterbodies from the National Hydrograpy Dataset (NHD) can be shown with varying level of detail and base maps can be included from the \CRANpkg{ggmap} package [@Kahle13].  NHD data were included to provide more specific information on hydrologic features of interest because of insufficient detail provided by standard basemaps.  These datasets are available from an external source and clipped to an approximate bounding box for the selected stations.  Currently, only flowlines and waterbodies that intersect Massachusetts are included, although examples are provided in a package [vignette](https://massbays-tech.github.io/MassWateR/articles/modifying.html#modifying-masswater-plots) that demonstrate how to include custom shapefiles in the map.  The below example shows mean dissolved oxygen concentrations using the medium level of detail for the NHD maps (`addwater = "medium"`).

@@

The last part of the \CRANpkg{MassWateR} workflow is preparation of data for submission to WQX.  A single function, `tabMWRwqx()`, is provided to format data inputs using a [template](https://massbays-tech.github.io/MassWateR/RESOURCES.html) designed for data submission.  This function will export a single Excel workbook with three sheets, named "Project", "Locations", and "Results", all of which are required by USEPA for data submission. The output is populated with as much content as possible based on information in the input files and the user can manually adjust the workbook prior to submission. All required columns are present, but individual rows must be verified for completeness and accuracy by the user before uploading the data.

The workbook can be created as follows by including the required files and specifying an output directory where the Excel file is saved. Once the function is done running, a message indicating success and where the file is located is returned.  Submitting the data to WQX simply requires uploading the output file into the data portal. After passing checks within WQX, the data are archived in the portal within 3-5 business days and made accessible to downstream users for public consumption.

@@

Additional templates and instructions for data submission are available on the [package website](https://massbays-tech.github.io/MassWateR/index.html).  It is assumed that users are familiar with the WQX data submission portal if they are already using the package, though [training materials](https://www.epa.gov/waterdata/learn-more-about-water-quality-data) from USEPA are available.  \CRANpkg{MassWateR} eliminates the need to format the data by hand and it is expected to increase the amount of data made available on the WQP (via WQX upload) as the user base for the package increases. Further, the authors worked with USEPA WQX staff to improve WQX handling of Quality Control-coded data, by improving the list of potential field names and highlighting the need to better connect paired data (e.g., to pair lab duplicates, we created activity types "Quality Control Lab Duplicate 1" and "Quality Control Lab Duplicate 2"). Custom WQX import configurations were also developed to streamline data submission, which were made available publicly to all WQX users.
